Measuring contrast processing in the visual system using SSVEP (and VEP?)

### 2024-04-01 review.qmd
Adding gain control words

### The development of contrast processing

An early use of the contrast response function was to provide an automated estimate of spatial contrast sensitivity in infants, without requiring behavioural responses.

In well-motivated adults, psychophysical measurements of contrast sensitivity remain the gold standard. However, it is difficult and time consuming to obtain reliable psychophysical data from infants. In these cases, SSVEP measurements represent a fast and efficient method for measuring low-level visual responses [Norcia, Tyler, Braddick].

Because SSVEP response at detection threshold are very small, this was achieved by measuring the contrast response function at relatively high levels, and extrapolating back along the function (either contrast vs response measured at a constant spatial frequency or spatial frequency vs response at a constant contrast level) to estimate its intercept with the x-axis. This contrast level was shown to correspond approximately with psychophysically measured detection thresholds [@DaleNorciaTyler86].

Measuring the SSVEP at multiple levels was made faster by the development of the ‘sweep VEP’ paradigm in which the stimulus changed its contrast, spatial frequency or some other property, throughout a trial [Tyler, Nakayama etc 1979]. To avoid hysteresis effects, the sweep is sometimes conducted both up and down in the same experiment [Tyler, Norcia etc: 1985, 1990 etc]. This general approach revealed much about the development of visual abilities in infants [@braddick, @atkinson]. In general, SSVEP measurements of infant vision have revealed that the development of spatial acuity for both achromatic and chromatic contrast as well as stereoscopic depth perception is faster than had been supposed previously based on behavioural readouts [@dobsonVisualAcuityHuman1978; @norciaSpatialFrequencySweep1985, ]. At least some of this difference is likely due to the high SNR of the SSVEP technique.

@@

The sweep VEP (really, a sweep 'SSVEP') technique is now commonly-used to obtain a rapid and objective measurement of visual acuity. In particular, because of its relative speed and simplicity, this technique has now become a standard for conducting tests of visual acuity in very young subjects or where behavioural tests are not appropriate [Ridder2004 Diva vs Enfant, Hoffman, Bach and Farmer].

# Contrast processing - gain control
Previous figures showed how a single-input SSVEP might generate a single peak in the FT response which, in turn, can be used as a read-out of neuronal activity.  Here, we show that adding a second contrast component can generate intermodulation and that the two components interact. The nature of the interaction is determined by the precise computations happening as the contrast signal moves from retina to cortex.

>>>>

### Measuring the development of contrast processing

An early use of the contrast response function was to provide an objective estimate of spatial contrast sensitivity in infants, without requiring behavioural responses.

In well-motivated adults, psychophysical measurements of contrast sensitivity remain the gold standard. However, it is difficult and time consuming to obtain reliable psychophysical data from infants. In these cases, SSVEP measurements represent a fast and efficient method for measuring low-level visual responses [Norcia, Tyler, Braddick] and the high SNR of SSVEP means that infants need only look at the screen for short periods of time.

Because SSVEP response at detection threshold are very small, estimating this threshold is achieved by measuring the contrast response function at relatively high levels, and extrapolating back along the function (either contrast vs response measured at a constant spatial frequency or spatial frequency vs response at a constant contrast level) to estimate its intercept with the x-axis. This contrast level was shown to correspond approximately with psychophysically measured detection thresholds [@DaleNorciaTyler86].

Measuring the SSVEP at multiple levels was made faster by the development of the ‘sweep VEP’ paradigm in which the stimulus changed its contrast, spatial frequency or some other property, throughout a trial [@Tyler, Nakayama etc 1979]. To avoid hysteresis effects, the sweep is sometimes conducted both up and down in the same experiment [Tyler, Norcia etc: 1985, 1990 etc].

This general approach has revealed much about the development of visual abilities in infants [@braddick, @atkinson]. In general, SSVEP measurements of infant vision have revealed that spatial acuity for both achromatic and chromatic contrast as well as stereoscopic depth perception develops earlier than had been supposed previously based on behavioural readouts [@dobsonVisualAcuityHuman1978; @norciaSpatialFrequencySweep1985, ] with both chromatic and achromatic contrast detection reaching near-adult levels by around six months. At least some of this difference is likely due to the relatively objectivity and high SNR of the SSVEP technique compared to other methods such as preferential looking which require careful measurement of the infant's gaze direction. One open question is how more complex contrast computations develop - and in particular how the human brain develop the tightly-tuned 'gain control' mechanisms that allow it to detect and discriminate contrast across very different visual scenes.

@@

The sweep VEP (really, a sweep 'SSVEP') technique is now commonly-used to obtain a rapid and objective measurement of visual acuity. In particular, because of its relative speed and simplicity, this technique has now become a standard for conducting tests of visual acuity in very young subjects or where behavioural tests are not appropriate [@Ridder2004 Diva vs Enfant, @Hoffman, @Bach and @Farmer].

# Contrast processing - linear and nonlinear

Neurons have a limited dynamic range yet they can transmit information about  visual stimuli that span many order of magnitude. In the domain of contrast, to some extent this is accomplished at a population level - individual neurons typically implement a non-linear,sigmoidal contrast vs response (CRF) transducer and different neurons exhibit peak sensitivity (the maximum slope of the function) at different contrast levels [@Early matteo stuff, Busse refs]. Individual neurons at multiple stages of the visual hierarchy also change their sensitivity depending on the average spatiotemporal contrast energy of their environment. This normalisation process is dynamic and nonlinear and is well-modeled by a hyperbolic ratio function in which the response of each neuron is modulated by a local 'gain pool' composed of the summed respnses of the local neuronal population.[@Heeger 92, @Carandini and Heeger 2011 @Busse 2008, @Baker and Wade]

Previous figures showed how a single-input SSVEP might generate a single set of peaks in the frequency domain response which, in turn, can be used as a read-out of neuronal activity.

Here, we show that adding a second contrast component can generate intermodulation and that the two components interact. The nature of the interaction is determined by the precise computations happening as the contrast signal moves from retina to cortex.
